A Continuation-Based Solution\break of the Linearity Challenge

### 2025-12-29 paper/Process.lagda.tex
tuning
global revision

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which builds a process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax: proving that a context $\Context$ satisfies the
predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is trivially enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable $X$, whereas $\Context$ is a
typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

>>>>

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which builds a process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax (HOAS): proving that a context $\Context$ satisfies
the predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$
of\Luca{Da definire} \ForallRule is trivially enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable $X$, whereas $\Context$ is a
typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

### 2025-12-29 paper/calculus.tex
global revision

In this section we give a cursory presentation of \Calculus starting from its
types (\Cref{sec:types}), which coincide with classical linear logic
propositions, then moving on to the syntax of processes (\Cref{sec:syntax}),
their reduction semantics (\Cref{sec:semantics}), the typing rules
(\Cref{sec:typing-rules}) and the formulation of the properties ensured by the
type system (\Cref{sec:properties}).

@@

The interpretation of linear logic propositions as behaviours is fairly
standard. Constants and connectives describe \emph{linear channels}, which must
be used for a single communication, whereas the modalities $\OfCourse$ and
$\WhyNot$ describe \emph{shared channels}.
%
The multiplicative constants $\Bot$ and $\One$ describe channels used for
receiving/sending an empty message (without continuations).
%
The additive connectives $A \With B$ and $A \Plus B$ describe  channels used for
receiving/sending either a continuation of type $A$ or a continuation of type
$B$. The sender selects one of the two possibilities, while the receiver offers
both.
%
The multiplicative connectives $A \Par B$ and $A \Ten B$ describe channels used
for receiving/sending two continuations, one of type $A$ and the other of type
$B$.
%
The additive constants can play the role of smallest/largest type in type
systems that support a notion of subtyping~\citep{HornePadovani24}. We will not
be using them in this work.
%
The quantifiers $\forall X.A$ and $\exists X.A$ describe channels used for
receiving/sending a type $X$ along with a continuation of type $A$. They are
useful to describe parametric protocol polymorphism.
%
Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
$\WhyNot A$ describe shared channels on which servers and clients accept and
request connections of type $A$.

The notions of \emph{free type variables}, of \emph{duality} and of \emph{type
substitution} are standard~\cite{Wadler14}. In particular, we write $\fv{A}$ for
the set of type variables occurring free in $A$, we write $\dual{A}$ for the
dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the free
occurrences of $X$ in $A$ with $B$.

@@

The syntax of processes makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that each
message sent on one of the channels is forwarded to the other. As discussed in
the literature~\cite{LindleyMorris15}, this form is useful for modeling
higher-order sessions, that is the exchange of an existing channel on another
channel.
%
The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
of an empty message on the channel $x$. The latter process has no continuation,
while the former process continues as $P$ once the message has been received.
%
The process $\Fail\x$ can be used to denote a failure concerning the channel
$x$.
%
The process $\Case\x\z{Q_1}{Q_2}$ offers a choice on channel $x$ and continues
as either $Q_1$ or $Q_2$ with $z$ bound to the received continuation depending
on which branch is selected.
%
The process $\Select\x{\InTag_i}\z.P$ performs a choice (represented by a label
$\InTag_i$ with $i=1,2$) and sends a fresh continuation $z$ on the channel $x$.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of two fresh continuations channels $y$ and $z$ on the channel $x$. The receiver
can use $y$ and $z$ in whatever order. The sender forks into $P$ and $Q$, each
using $y$ and $z$ respectively. These forms are often used in conjunction with
links to model higher-order sessions.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

@@

The notions of free and bound channels are fairly standard, bearing in mind that
output prefixes bind continuation channels in (some) continuation processes. In
particular, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while
$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
$P$.
%
We write $\fn{P}$ for the set of channels occurring free in $P$ and we identify
processes up to renaming of bound channels.

\begin{example}
    \label{ex:booleans}
    We can represent the protocol of a boolean value being produced as the type
    $\Bool \eqdef \One \Plus \One$ and that of a boolean value being consumed as
    its dual $\dual\Bool = \Bot \With \Bot$. Following these types, the boolean
    constants can be modeled by the processes
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{x'}.\Close{x'}
        \qquad
        \FalseP(x) \eqdef \Select\x\RightTag{x'}.\Close{x'}
    \]
    and the boolean negation function by the process
    \[
        \NotP(x,y) \eqdef
        \BigCase\x{x'}{
            \Wait{x'}.\FalseP(y)
        }{
            \Wait{x'}.\TrueP(y)
        }
    \]

@@

The operational semantics of \Calculus is shown in~\Cref{tab:semantics} and is
given by two relations: a \emph{structural pre-congruence} relation $\pcong$ and
a \emph{reduction} relation $\red$. Structural pre-congruence relates processes
that we consider to be essentially indistinguishable. Reduction relates
processes in which a communication occurs. Let us describe the two relations
more in detail.

@@

There is no pre-congruence rule asserting that parallel composition is
associative. While this is a sound property of parallel composition, it turns
out to be unnecessary for proving the properties of \Calculus that we consider. We
leave it out in light of the forthcoming formalisation of the calculus, in which
every simplification is welcome, although standard presentations of classical
linear logic admit such law~\cite{Wadler14,Doumane17}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

Reduction is defined by the \rulename{r-*} rules.
%
The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists and \RConnect
are in one-to-one correspondence with the well-known reductions of session
calculi and the principal cut reductions of linear logic. These reductions erase
the topmost cut and replace it with zero, one or two new cuts, depending on the
number of continuations that are exchanged. Note that \RLink eliminates a link
$\Link\x\y$ by substituting $y$ for $x$ in the scope of $x$ and \RClose models
the communication of an empty message (without continuations), hence no new cut
is present in the reduct. \RSelect, \RFork and \RConnect are standard and
\RExists models the instantiation of a polymorphic variable $X$ as the
communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
replacing every free occurrence of the type variable $X$ with $A$.
%
The reduction rules \RWeaken and \RContract respectively represent the disposal
of an unused server and the duplication of a server. In these rules we use some
convenient (albeit slightly informal) notation for denoting sequences of
(pairwise distinct) channels and prefixes. In particular, $\seqof\z$ stands for
a sequence $z_1, \dots, z_n$ of channels, $\Weaken{\seqof\z}.Q$ stands for
$\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

@@

\begin{example}
    \label{ex:neg}
    %
    We illustrate an example of higher-order session by modeling an echo server
    that consumes boolean values and sends them back unchanged. The protocol
    implemented by the server is described by the type $\OfCourse(\dual\Bool
    \Par (\Bool \Ten \One))$ where the modality~$\OfCourse$ indicates that the
    server is able to accept an unbounded number of requests and the type
    $\dual\Bool \Par (\Bool \Ten \One)$ describes the sequence of actions
    performed by the server at each connection with a client: the server first
    consumes a boolean value (say $u$, of type $\dual\Bool$), then produces
    another boolean value (say $v$, of type $\Bool$) and finally sends an empty
    message. We can model the server and a possible client thus:
    \begin{align*}
        \ServerP(x) & \eqdef
        \Server\x\y.
        \Join\y\u{y'}.
        \Fork{y'}\v{y''}{
            \Link\v\u
        }{
            \Close{y''}
        } \\
        \ClientP(x,z) & \eqdef
        \Client\x\y.
        \Fork\y\u{y'}{
            \TrueP(u)
        }{
            \Join{y'}\v{y''}.
            \Wait{y''}.
            \Link\v\z
        }
    \end{align*}

    Notice the use of continuations for chaining communications together. In the
    server, the link $\Link\v\u$ merges $v$ and $u$ so as to send the same
    channel $u$ received from the client. In the client, the link $\Link\v\z$
    ``assigns'' the message $v$ received from the server to the free channel
    $z$, which represents the result of the interaction.
    %
    If we write $\reds$ for the reflexive, transitive closure of $\red$ it is
    easy to verify that $\Cut\x{\ClientP(x)}{\ServerP(x)} \red^* \TrueP(z)$,
    as expected.
    %
    \eoe
\end{example}

@@

We use typing contexts (\ie sequents) to keep track of the type of channels.
Typing contexts are finite maps from channels to types written as $x_1 : A_1,
\dots, x_n : A_n$ and ranged over by $\ContextC$ and $\ContextD$. We write
$\ContextC, \ContextD$ for the union of $\ContextC$ and $\ContextD$ when they
have disjoint domains. We write $\WhyNot\Context$ for some context $\Context =
x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in its range
are prefixed by the modality~$\WhyNot$. We call these types \emph{unrestricted}
because they are used to denote shared channels that can be (explicitly)
discarded and duplicated.

\begin{table}[t]
    \caption{Typing rules of \Calculus.}
    \label{tab:typing-rules}
    \begin{tabular}{@{}c@{}}
        \begin{mathpar}
            \inferrule[\LinkRule]{~}{
                \wtp{\Link\x\y}{x : A, y : \dual{A}}
            }
            \and
            \inferrule[\FailRule]{~}{
                \wtp{\Fail\x}{x : \Top, \Context}
            }
            \and
            \inferrule[\WaitRule]{
                \wtp{P}{\Context}
            }{
                \wtp{\Wait\x.P}{x : \Bot, \Context}
            }
            \and
            \inferrule[\CloseRule]{~}{
                \wtp{\Close\x}{x : \One}
            }
            \and
            \inferrule[\CaseRule]{
                \wtp{P}{y : A, \Context}
                \\
                \wtp{Q}{y : B, \Context}
            }{
                \wtp{\Case\x\y{P}{Q}}{x : A \With B, \Context}
            }
            \and
            \inferrule[\SelectRule]{
                \wtp{P}{y : A_i, \Context}
            }{
                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
            }
            \and
            \inferrule[\JoinRule]{
                \wtp{P}{y : A, z : B, \Context}
            }{
                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
            }
            \and
            \inferrule[\ForkRule]{
                \wtp{P}{y : A, \ContextC}
                \\
                \wtp{Q}{z : B, \ContextD}
            }{
                \wtp{\Fork\x\y\z{P}{Q}}{x : A \Ten B, \ContextC, \ContextD}
            }
            \and
            \inferrule[\ExistsRule]{
                \wtp{P}{y : B\subst{A}\X, \Context}
            }{
                \wtp{\Exists\x{A}\y.P}{x : \exists\X.B, \Context}
            }
            \and
            \inferrule[\ForallRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Forall\x\X\y.P}{x : \forall\X.A, \Context}
            }
            ~X\not\in\fn\Context
            \and
            \inferrule[\ServerRule]{
                \wtp{P}{y : A, \WhyNot\Context}
            }{
                \wtp{\Server\x\y.P}{x : \OfCourse A, \WhyNot\Context}
            }
            \and
            \inferrule[\ClientRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Client\x\y.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\WeakenRule]{
                \wtp{P}\Context
            }{
                \wtp{\Weaken\x.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\ContractRule]{
                \wtp{P}{y : \WhyNot A, z : \WhyNot A, \Context}
            }{
                \wtp{\Contract\x\y\z.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\CutRule]{
                \wtp{P}{x : A, \ContextC}
                \\
                \wtp{Q}{x : \dual{A}, \ContextD}
            }{
                \wtp{\Cut[A]\x{P}{Q}}{\ContextC, \ContextD}
            }
        \end{mathpar}
    \end{tabular}
\end{table}

Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
well typed in the context $\Context$. Equivalently, the judgment indicates that
the sequent $\wtp{}\Context$ is derivable and $P$ is a proof term corresponding
to the derivation for $\wtp{}\Context$.
%
The typing rules are shown in \Cref{tab:typing-rules}. They are in one-to-one
correspondence with -- and have exactly the same structure of -- the proof rules
of classical linear logic.
%
The reader may refer to the standard literature on propositions as
sessions~\citep{Wadler14,LindleyMorris15,GayVasconcelos25} for the
interpretation of the rules. The only relevant difference with the typing rules
in these papers is that the premises mentioning the continuation channel $z$
actually refer to the same channel $x$ on which the process in the conclusion is
acting.

@@

In order to formulate deadlock freedom, we first need to introduce some
terminology for referring to processes that are unable to make any progress. A
simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$. This
process is unable to reduce (because a process $\Close\x$ is meant to interact
with a process of the form $\Wait\x.P$) and, more generally, it is unable to
interact regardless of the context in which it occurs because the sub-processes
it contains are blocked on the channel $x$ that is restricted by the cut.
%
In general, the property of being a deadlock is not simply the inability to
reduce: there are irreducible processes that are not a deadlock because they
would be able to reduce if put into a suitable context. For example, $\Wait\x.P$
does not reduce, and yet it is not a deadlock because it would be able to make
progress when composed in parallel with $\Close\x$. Even a process like
$\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a deadlock,
because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using \SWait and
possibly \SComm. Let us make all this more precise.

>>>>

In this section we give a cursory presentation of \Calculus starting from its
types (\Cref{sec:types}) then moving on to the syntax of processes
(\Cref{sec:syntax}), their reduction semantics (\Cref{sec:semantics}), the
typing rules (\Cref{sec:typing-rules}) and the formulation of the properties
ensured by the type system (\Cref{sec:properties}).

@@

The interpretation of linear logic propositions as behaviours is fairly
standard. Constants and connectives describe \emph{linear channels}, which must
be used for a single communication, whereas the modalities $\OfCourse$ and
$\WhyNot$ describe \emph{shared channels}.
%
The multiplicative constants $\Bot$ and $\One$ describe channels used for
receiving/sending an empty message (without continuations).
%
The additive constants describe unusable channels. They can play the role of
smallest/largest type in type systems that support a notion of
subtyping~\citep{HornePadovani24}, but will mostly ignore them in this work.
%
The additive connectives $A \With B$ and $A \Plus B$ describe  channels used for
receiving/sending either a continuation of type $A$ or a continuation of type
$B$. The sender selects one of the two possibilities, while the receiver offers
both.
%
The multiplicative connectives $A \Par B$ and $A \Ten B$ describe channels used
for receiving/sending two continuations, one of type $A$ and the other of type
$B$.
%
The quantifiers $\forall X.A$ and $\exists X.A$ describe channels used for
receiving/sending a type $X$ along with a continuation of type $A$. They are
useful to describe parametric protocol polymorphism.
%
Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
$\WhyNot A$ describe shared channels on which servers and clients accept and
request connections of type $A$.

The notions of \emph{free type variables}, of \emph{duality} and of \emph{type
substitution} are standard~\cite{Wadler14}. In particular, we write $\dual{A}$
for the dual of $A$ and $A\subst{B}X$ for the type obtained by replacing the
free occurrences of $X$ in $A$ with $B$.

@@

The syntax of processes makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that each
message sent on one of the channels is forwarded to the other. As discussed in
the literature~\cite{LindleyMorris15} and illustrated in \Cref{ex:neg}, this
form is useful for modeling the exchange of an existing channel on another
channel.
%
The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
of an empty message on the channel $x$. The latter process terminates after the
message has been sent, while the former process continues as $P$ once the
message has been received.
%
The process $\Fail\x$ can be used to denote a failure concerning the channel
$x$.
%
The process $\Case\x\z{Q_1}{Q_2}$ offers a choice on channel $x$ and continues
as either $Q_1$ or $Q_2$ depending on which branch is selected with $z$ bound to
the received continuation channel.
%
The process $\Select\x{\InTag_i}\z.P$ performs a choice (represented by a label
$\InTag_i$ with $i=1,2$) and sends a fresh continuation channel $z$ on the
channel $x$.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of two fresh continuations channels $y$ and $z$ on the channel $x$. The receiver
can use $y$ and $z$ in whatever order. The sender forks into $P$ and $Q$, each
using $y$ and $z$ respectively.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

@@

The notions of free and bound channels are fairly standard, bearing in mind that
output prefixes bind continuation channels in (some) continuation processes. For
instance, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while $\Fork\x\y\z{P}{Q}$
binds $y$ in $P$ but not in $Q$ and binds $z$ in $Q$ but not in $P$.
%
We write $\fn{P}$ for the set of channels occurring free in $P$ and we identify
processes up to renaming of bound channels.

\begin{example}
    \label{ex:booleans}
    We can represent the protocol of a boolean value being produced as the type
    $\Bool \eqdef \One \Plus \One$ and that of a boolean value being consumed as
    its dual $\dual\Bool = \Bot \With \Bot$. Following these types, the boolean
    constants can be modeled by the processes
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{x'}.\Close{x'}
        \qquad
        \FalseP(x) \eqdef \Select\x\RightTag{x'}.\Close{x'}
    \]
    and the boolean negation function by the process
    \[
        \NotP(x,y) \eqdef
        \Case\x{x'}{
            \Wait{x'}.\FalseP(y)
        }{
            \Wait{x'}.\TrueP(y)
        }
    \]

@@

The operational semantics of \Calculus is shown in~\Cref{tab:semantics} and is
given by two relations: a \emph{structural pre-congruence} relation $\pcong$,
which relates essentially indistinguishable processes, and a \emph{reduction}
relation $\red$, which models communications. Let us describe the two relations
more in detail.

@@

There is no pre-congruence rule asserting that parallel composition is
associative. While this is a sound property of parallel composition, it turns
out to be unnecessary for proving the properties of \Calculus that we consider. We
leave it out in light of the forthcoming formalisation of the calculus, in which
every simplification is welcome.
%
Another rule that deserves attention is \SServer. In this case, a cut can be
pushed underneath a server prefix $\Server\y\u$ only provided that the other
process in the cut is also a server on the channel restricted by the cut.

Reduction is defined by the \rulename{r-*} rules, most of them coinciding with
principal cut reductions of linear logic.
%
The rules \RLink, \RClose, \RSelect, \RFork and \RExists erase the topmost cut
and replace it with zero, one or two new cuts, depending on the number of
continuation channels that are exchanged.
%
Note that the rule \RLink eliminates a link $\Link\x\y$ by substituting $y$ for
$x$ in the scope of $x$ and \RClose models the communication of an empty message
(without continuations). The rule \RExists models the instantiation of a
polymorphic variable $X$ as the communication of a type $A$. We write
$Q\subst{A}\X$ for the process obtained by replacing every free occurrence of
the type variable $X$ with $A$.
%
The rule \RConnect models the connection between a client and a server, whereas
the rules \RWeaken and \RContract respectively model the disposal of an unused
server and the duplication of a server. In these rules we use some slightly
informal notation for denoting sequences of (pairwise distinct) channels and
prefixes. In particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of
channels, $\Weaken{\seqof\z}.Q$ stands for a sequence
$\Weaken{z_1}\dots\Weaken{z_n}.Q$ of weakening prefixes and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for a sequence
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$ of contraction
prefixes.

@@

\begin{example}
    \label{ex:neg}
    %
    In this example we illustrate the role of links for the communication of
    free channels by modeling an echo server that consumes boolean values and
    sends them back unchanged.
    %
    The protocol of the server we want to model is described by the type
    $\OfCourse(\dual\Bool \Par (\Bool \Ten \One))$ where the
    modality~$\OfCourse$ indicates that the server is able to accept an
    unbounded number of requests and the type $\dual\Bool \Par (\Bool \Ten
    \One)$ describes the sequence of actions performed by the server at each
    connection with a client: the server first consumes a boolean value (say
    $u$, of type $\dual\Bool$), then produces another boolean value (say $v$, of
    type $\Bool$) and finally sends an empty message. We can model the server
    and a possible client thus:
    \begin{align*}
        \ServerP(x) & \eqdef
        \Server\x\y.
        \Join\y\u{y'}.
        \Fork{y'}\v{y''}{
            \Link\v\u
        }{
            \Close{y''}
        } \\
        \ClientP(x,z) & \eqdef
        \Client\x\y.
        \Fork\y\u{y'}{
            \TrueP(u)
        }{
            \Join{y'}\v{y''}.
            \Wait{y''}.
            \Link\v\z
        }
    \end{align*}

    Notice the use of continuations for chaining communications together. In the
    server, the link $\Link\v\u$ merges $v$ and $u$ so as to send the same
    channel $u$ received from the client. In the client, the link $\Link\v\z$
    ``assigns'' the message $v$ received from the server to the free channel
    $z$, which represents the result of the interaction.
    %
    If we write $\reds$ for the reflexive, transitive closure of $\red$ it is
    easy to verify that $\Cut\x{\ClientP(x)}{\ServerP(x)} \red^* \TrueP(z)$.
    %
    \eoe
\end{example}

@@

We use typing contexts (\ie sequents) to keep track of the type of channels in
processes. Typing contexts are finite maps from channels to types written as
$x_1 : A_1, \dots, x_n : A_n$ and ranged over by $\ContextC$ and $\ContextD$. We
write $\ContextC, \ContextD$ for the union of $\ContextC$ and $\ContextD$ when
they have disjoint domains. We write $\WhyNot\Context$ for some context
$\Context = x_1 : \WhyNot A_1, \dots, x_n : \WhyNot A_n$ where all the types in
its range are prefixed by the modality~$\WhyNot$. We call these types
\emph{unrestricted} because they are used to denote shared channels that can be
(explicitly) discarded and duplicated.

\begin{table}[t]
    \caption{Typing rules of \Calculus.}
    \label{tab:typing-rules}
    \begin{tabular}{@{}c@{}}
        \begin{mathpar}
            \inferrule[\LinkRule]{~}{
                \wtp{\Link\x\y}{x : A, y : \dual{A}}
            }
            \and
            \inferrule[\FailRule]{~}{
                \wtp{\Fail\x}{x : \Top, \Context}
            }
            \and
            \inferrule[\WaitRule]{
                \wtp{P}{\Context}
            }{
                \wtp{\Wait\x.P}{x : \Bot, \Context}
            }
            \and
            \inferrule[\CloseRule]{~}{
                \wtp{\Close\x}{x : \One}
            }
            \and
            \inferrule[\CaseRule]{
                \wtp{P}{y : A, \Context}
                \\
                \wtp{Q}{y : B, \Context}
            }{
                \wtp{\Case\x\y{P}{Q}}{x : A \With B, \Context}
            }
            \and
            \inferrule[\SelectRule]{
                \wtp{P}{y : A_i, \Context}
            }{
                \wtp{\Select\x{\InTag_i}\y.P}{x : A_1 \Plus A_2, \Context}
            }
            \and
            \inferrule[\JoinRule]{
                \wtp{P}{y : A, z : B, \Context}
            }{
                \wtp{\Join\x\y\z.P}{x : A \Par B, \Context}
            }
            \and
            \inferrule[\ForkRule]{
                \wtp{P}{y : A, \ContextC}
                \\
                \wtp{Q}{z : B, \ContextD}
            }{
                \wtp{\Fork\x\y\z{P}{Q}}{x : A \Ten B, \ContextC, \ContextD}
            }
            \and
            \inferrule[\ExistsRule]{
                \wtp{P}{y : B\subst{A}\X, \Context}
            }{
                \wtp{\Exists\x{A}\y.P}{x : \exists\X.B, \Context}
            }
            \and
            \inferrule[\ForallRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Forall\x\X\y.P}{x : \forall\X.A, \Context}
            }
            ~X\not\in\fv\Context
            \and
            \inferrule[\ServerRule]{
                \wtp{P}{y : A, \WhyNot\Context}
            }{
                \wtp{\Server\x\y.P}{x : \OfCourse A, \WhyNot\Context}
            }
            \and
            \inferrule[\ClientRule]{
                \wtp{P}{y : A, \Context}
            }{
                \wtp{\Client\x\y.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\WeakenRule]{
                \wtp{P}\Context
            }{
                \wtp{\Weaken\x.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\ContractRule]{
                \wtp{P}{y : \WhyNot A, z : \WhyNot A, \Context}
            }{
                \wtp{\Contract\x\y\z.P}{x : \WhyNot A, \Context}
            }
            \and
            \inferrule[\CutRule]{
                \wtp{P}{x : A, \ContextC}
                \\
                \wtp{Q}{x : \dual{A}, \ContextD}
            }{
                \wtp{\Cut[A]\x{P}{Q}}{\ContextC, \ContextD}
            }
        \end{mathpar}
    \end{tabular}
\end{table}

Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
well typed in the context $\Context$. Equivalently, the judgment indicates that
the sequent $\wtp{}\Context$ is derivable and $P$ is a proof term corresponding
to the derivation for $\wtp{}\Context$.
%
The typing rules are shown in \Cref{tab:typing-rules}. They are in one-to-one
correspondence with -- and have exactly the same structure of -- the proof rules
of classical linear logic.
%
The reader may refer to the standard literature on propositions as
sessions~\citep{Wadler14,LindleyMorris15,GayVasconcelos25} for the
interpretation of the rules. The only relevant difference with our typing rules
is that the premises mentioning the continuation channel $z$ actually refer to
the same channel $x$ on which the process in the conclusion is acting.
%
The side condition $X \not\in \fv\Context$ in the rule \ExistsRule checks that
the type variable $X$ does not occur free in the types of $\Context$ and
therefore can be generalised.

@@

In order to formulate deadlock freedom, we first need to introduce some
terminology for referring to processes that are unable to make any progress. A
simple example of deadlock is the process $\Cut\x{\Close\x}{\Close\x}$. This
process is unable to reduce (because a process $\Close\x$ is meant to interact
with a process of the form $\Wait\x.P$) and, more generally, it is unable to
interact regardless of the context in which it is placed because the
sub-processes it contains are blocked on the channel $x$ that is restricted by
the cut.
%
In general, the property of being a deadlock is not simply the inability to
reduce: there are irreducible processes that are not a deadlock because they
would be able to reduce if put into a suitable context. For example, $\Wait\x.P$
does not reduce, and yet it is not a deadlock because it would be able to make
progress when composed in parallel with $\Close\x$. Even a process like
$\Cut\y{\Wait\y.P}{\Wait\z.Q}$ where $x\ne y,z$ cannot be considered a deadlock,
because the prefixes $\Wait\y$ and $\Wait\z$ could be exposed using \SWait and
possibly \SComm. Let us make all this more precise.
